---
title: "OptSLDP: An Optimized Selective Linkage Disequilibrium Pruning Pipeline"
author: "Félicien Akohoue"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{OptSLDP: An Optimized Selective Linkage Disequilibrium Pruning Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  warning    = FALSE,
  message    = FALSE,
  error      = TRUE
)
```

---

# Introduction

`OptSLDP` is an **optimized and extended** R implementation of the Selective
Linkage Disequilibrium (LD) Pruning pipeline for genomic prediction panel
construction. The package builds on the algorithm of
Zhu et al. (2023), which demonstrated that preserving markers in strong LD
with statistically important SNPs substantially improves genomic prediction
accuracy compared to standard uniform LD pruning. `OptSLDP` extends that
work with four concrete algorithmic improvements described in detail below.

## Why standard LD pruning is not enough for genomic prediction

Standard LD pruning (e.g., PLINK `--indep-pairwise`) removes one marker from
every highly correlated pair without regard to whether either marker is
associated with the trait of interest. When a QTL region contains a cluster of
correlated markers, all but one are discarded — yet the retained singleton
may not be the best proxy for the causal variant, and statistical evidence
about the region is reduced.

SLDP addresses this by partitioning the marker set into two groups before
pruning:

- **Important SNPs** ($\mathcal{I}$): SNPs that are statistically associated
  with at least one trait, plus all markers in strong LD with them within a
  positional window. These are retained in full.
- **Background SNPs** ($\mathcal{B} = \Omega \setminus \mathcal{I}$): the
  remainder, which receives standard greedy LD pruning to control redundancy.

The final panel is $\mathcal{I} \cup \text{retained}(\mathcal{B})$.

## How `OptSLDP` improves on the original algorithm

| Improvement | Description |
|-------------|-------------|
| Chunked reading | Genotype files > 200 K rows read in 50 K-row chunks with a pre-allocated matrix; peak RAM is one chunk rather than twice the full file |
| Per-chromosome `gc()` | `gc(FALSE)` called after each chromosome's LD matrix is released in both the pre-pruning and background-pruning loops |
| Multi-trait union protection | Screening and candidate selection run independently per trait; the union of all per-trait candidate sets drives expansion and protection |
| Covariate-adjusted screening | Phenotypes are residualised on shared covariates once before the SNP scan, equivalent to fitting covariates in every marginal regression |
| Vectorised OLS screening | Matrix algebra (`tcrossprod` via BLAS DGEMM) replaces per-SNP `lm()` calls; 50-200x faster for large panels |
| C++ LD kernel | `r2_subset_cpp()`, `above_threshold_subset_cpp()`, `greedy_prune_r2_cpp()` compiled via RcppArmadillo; zero R interpreter overhead |
| Batched LD expansion | Genotypes extracted once per chromosome; `foverlaps()` O(n log n) joins replace O(n^2) loops; GDS reads reduced from thousands to ~11 |

---

# Installation

```{r install, eval = FALSE}
# Install from GitHub (with vignettes)
install.packages("remotes")
remotes::install_github("FAkohoue/OptSLDP",
  build_vignettes = TRUE,
  dependencies    = TRUE
)

# Optional: VCF support (Bioconductor)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("VariantAnnotation", "GenomeInfoDb",
                       "SummarizedExperiment", "Biostrings", "S4Vectors"))

# Optional: GDS backend for > 2 M SNP panels (Bioconductor)
BiocManager::install(c("SNPRelate", "gdsfmt"))
```

Load the package:

```{r library}
library(OptSLDP)
```

---

# Example data

`OptSLDP` ships with a small synthetic dataset that can be used to run all
examples in this vignette without any external files. The dataset was
generated with a known genetic architecture so that results can be validated
against ground truth.

## Dataset design

The example data contains:

- **40 SNPs** on 2 chromosomes (20 per chromosome), positions 10 000–200 000
- **50 samples** (Line01–Line50)
- **Three LD blocks**:
  - Block A: SNP001–SNP008 (chr1, $r^2 \approx 0.80$)
  - Block B: SNP012–SNP018 (chr1, $r^2 \approx 0.60$)
  - Block C: SNP023–SNP030 (chr2, $r^2 \approx 0.80$)
- **~5% missing genotype values** encoded as `NA` (numeric), `NN` (HapMap),
  or `./.` (VCF)
- **Two traits with independent genetic architectures**:

| Trait | QTN | Effects | Heritability | Notes |
|-------|-----|---------|--------------|-------|
| Trait1 | SNP003, SNP015, SNP025 | +0.80, −0.50, +0.65 | $h^2 = 0.55$ | One QTN per LD block |
| Trait2 | SNP006, SNP028, SNP020 | +0.70, −0.60, +0.45 | $h^2 = 0.45$ | SNP020 is a singleton outside all LD blocks |

The partial overlap of Trait1 and Trait2 across LD blocks (A and C), combined
with Trait2's unique singleton QTN at SNP020, tests whether the union-protection
logic correctly retains SNP020 even though it is never a candidate under Trait1
alone.

## Accessing example files

```{r paths}
geno_file  <- system.file("extdata", "example_genotypes_numeric.csv",
                           package = "OptSLDP")
pheno_file <- system.file("extdata", "example_phenotype.csv",
                           package = "OptSLDP")
hmp_file   <- system.file("extdata", "example_genotypes.hmp.txt",
                           package = "OptSLDP")
vcf_file   <- system.file("extdata", "example_genotypes.vcf",
                           package = "OptSLDP")
```

## Inspecting the example files

```{r inspect_data}
# Phenotype file: Sample, Trait1, Trait2, PC1, PC2
pheno_df <- read.csv(pheno_file)
head(pheno_df)
dim(pheno_df)

# First few rows of the genotype file
geno_head <- read.csv(geno_file, check.names = FALSE)
geno_head[1:5, 1:8]
```

---

# Reading input data

## Genotype formats

`OptSLDP` supports three genotype input formats detected automatically from
the file extension.

### Numeric dosage (CSV / TXT)

The standard format has columns `SNP, CHR, POS, REF, ALT` followed by one
column per sample. Values must be in $\{0, 1, 2, \text{NA}\}$.

```{r read_numeric}
geno_obj <- read_numeric_genotype(geno_file)

# Structure of the returned list
str(geno_obj, max.level = 1)

# SNP metadata
head(geno_obj$snp_info)

# Dimensions of the genotype matrix (SNPs × samples)
dim(geno_obj$geno_mat)

# Sample IDs
head(geno_obj$sample_ids)
```

For files exceeding 200 000 rows, the reader automatically switches to a
two-pass chunked strategy:

1. **Pass 1** — header-only `fread(nrows = 0)` to determine column names and
   count rows.
2. **Pre-allocation** — `matrix(NA_real_, nrow, ncol)` allocated once.
3. **Pass 2** — data filled in 50 000-row chunks; each chunk is released and
   `gc(FALSE)` called before reading the next.

The threshold and chunk size are configurable:

```{r read_numeric_chunks, eval = FALSE}
geno_obj <- read_numeric_genotype(
  file            = "large_panel.csv",
  chunk_rows      = 50000L,    # rows per chunk (default)
  chunk_threshold = 200000L    # rows above which chunked reading is used
)
```

### HapMap format

```{r read_hapmap}
hmp_obj <- read_hapmap_genotype(hmp_file)
str(hmp_obj, max.level = 1)
head(hmp_obj$snp_info)
```

HapMap nucleotide calls (`"AA"`, `"AT"`, `"TT"`, `"NN"`) are converted to
additive dosage (0/1/2/NA) relative to the REF allele inferred from the
`alleles` column.

### VCF format

```{r read_vcf, eval = requireNamespace("VariantAnnotation", quietly = TRUE)}
vcf_obj <- read_vcf_genotype(vcf_file)
str(vcf_obj, max.level = 1)
```

Both phased (`0|1`) and unphased (`0/1`) GT fields are accepted. The first 10
SNPs in the example VCF use phased notation; the remainder use unphased — this
exercises both parsing branches. Multi-allelic sites use the first ALT allele.

### Automatic format dispatch

```{r read_auto}
# Format detected from extension: .csv → numeric dosage
geno_auto <- read_genotype(geno_file)
identical(geno_auto$format, "numeric")

# Chromosome names: "chr1" and "1" are normalised at read time
# so downstream matching always works
head(geno_auto$snp_info$CHR)
```

## Phenotype file

```{r read_pheno}
pheno_obj <- read_phenotype(
  file         = pheno_file,
  sample_col   = "Sample",
  trait_col    = "Trait1",
  covar_cols   = c("PC1", "PC2"),
  sample_order = geno_obj$sample_ids  # drives row reordering
)

# Single trait: returns a plain numeric vector
class(pheno_obj$phenotype)
length(pheno_obj$phenotype)

# Covariates as a data.frame
head(pheno_obj$covariates)

# Sample IDs in the aligned order
identical(pheno_obj$sample_ids, geno_obj$sample_ids)
```

For multiple traits, pass a character vector to `trait_col`. The return value
is then a named numeric matrix (samples × traits):

```{r read_pheno_multi}
pheno_multi <- read_phenotype(
  file         = pheno_file,
  sample_col   = "Sample",
  trait_col    = c("Trait1", "Trait2"),
  covar_cols   = c("PC1", "PC2"),
  sample_order = geno_obj$sample_ids
)

# Multiple traits: returns a matrix
class(pheno_multi$phenotype)
dim(pheno_multi$phenotype)
colnames(pheno_multi$phenotype)
```

---

# Quality control

## Minor allele frequency

The ALT allele frequency is estimated from the dosage matrix:

$$\text{AF}_i = \frac{\sum_j g_{ij}}{2\, n_i}$$

where $g_{ij} \in \{0, 1, 2, \text{NA}\}$ is the dosage for SNP $i$ in sample
$j$, and $n_i$ is the number of non-missing samples for SNP $i$. The minor
allele frequency is:

$$\text{MAF}_i = \min(\text{AF}_i,\; 1 - \text{AF}_i)$$

```{r maf}
maf_dt <- compute_maf(geno_obj$geno_mat)
head(maf_dt)

# Distribution of MAF across the example panel
summary(maf_dt$MAF)
```

```{r maf_plot, fig.cap = "MAF distribution across the 40 example SNPs."}
hist(
  maf_dt$MAF,
  breaks = 10,
  col    = "steelblue",
  border = "white",
  main   = "MAF distribution",
  xlab   = "Minor allele frequency",
  ylab   = "Number of SNPs"
)
abline(v = 0.05, col = "firebrick", lty = 2, lwd = 2)
legend("topright", legend = "MAF = 0.05 threshold",
       col = "firebrick", lty = 2, lwd = 2, bty = "n")
```

## MAF filtering

```{r maf_filter}
maf_res <- filter_snps_by_maf(
  snp_info      = geno_obj$snp_info,
  geno_mat      = geno_obj$geno_mat,
  maf_threshold = 0.05
)

cat("SNPs before MAF filter:", nrow(geno_obj$snp_info), "\n")
cat("SNPs after  MAF filter:", nrow(maf_res$snp_info),  "\n")
cat("Removed:", nrow(geno_obj$snp_info) - nrow(maf_res$snp_info), "\n")
```

## High-LD pre-pruning

Before screening, near-duplicate SNPs ($r^2 \ge \tau_{\text{pre}}$, default
0.99) are removed chromosome-by-chromosome with a greedy forward-selection
algorithm. This step targets perfect or near-perfect redundancy (duplicate
probes, collapsed haplotypes) without removing biological LD signal.

```{r preprune}
pre_res <- preprune_high_ld(
  snp_info     = maf_res$snp_info,
  geno_mat     = maf_res$geno_mat,
  r2_pre       = 0.99,
  verbose      = FALSE
)

cat("SNPs before pre-pruning:", nrow(maf_res$snp_info), "\n")
cat("SNPs after  pre-pruning:", nrow(pre_res$snp_info), "\n")
```

At the end of each chromosome's loop, the $r^2$ matrix is freed immediately
and `gc(FALSE)` is called, preventing heap fragmentation from accumulating
across chromosomes:

```r
# Internal logic in preprune_high_ld() and prune_background_snps()
r2_mat <- compute_r2_subset(geno_mat, chr_snps)
# ... pruning loop ...
rm(r2_mat, kept)
gc(FALSE)   # explicit release before the next chromosome
```

---

# Marginal SNP screening

## Statistical model

For each SNP $i$, a marginal linear regression is fitted:

$$y^* = \alpha + \beta_i\, g_i + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2 I)$$

where $y^*$ is the working phenotype (see covariate adjustment below), $g_i$
is the dosage vector for SNP $i$, and $\alpha$, $\beta_i$ are intercept and
effect size.

**Covariate adjustment.** When covariates $X_c$ are supplied, the phenotype
is residualised once before the SNP loop:

$$y^* = M_{X_c}\, y = \left[I - X_c(X_c^\top X_c)^{-1} X_c^\top\right] y$$

This is equivalent to fitting $y \sim X_c + g_i$ for every SNP but avoids
refitting covariates in each regression, giving an $O(n)$ saving over $O(n \cdot p)$.

**Reported statistics per SNP:**

| Column | Formula |
|--------|---------|
| `beta` | $\hat\beta_i = \text{Cov}(g_i, y^*) / \text{Var}(g_i)$ |
| `SE` | $\widehat{\text{SE}}(\hat\beta_i) = \sqrt{\hat\sigma^2 / \sum_j(g_{ij} - \bar g_i)^2}$, where $\hat\sigma^2 = \text{RSS}/(n-2)$ |
| `z_score` | $z_i = \hat\beta_i / \widehat{\text{SE}}(\hat\beta_i)$ |
| `P.value` | $p_i = 2\Pr(T_{n-2} > |z_i|)$ (two-tailed $t$-test) |
| `PVE` | $R^2_i$ from the marginal regression |
| `AF` | ALT allele frequency $= \bar g_i / 2$ |
| `MAF` | $\min(\text{AF}_i,\, 1 - \text{AF}_i)$ |

SNPs with fewer than 5 complete observations or zero variance receive `NA`
for all statistical columns.

## Single-trait screening

```{r screen_single}
geno_mat_filt <- pre_res$geno_mat
snp_info_filt <- pre_res$snp_info
phenotype_vec <- pheno_obj$phenotype
covariates_df <- pheno_obj$covariates

screen_stats <- compute_screening_stats(
  geno_mat = geno_mat_filt,
  y        = phenotype_vec,
  covar    = covariates_df,
  verbose  = FALSE
)

# Returns a single data.table
class(screen_stats)
head(screen_stats[order(screen_stats$P.value), ])
```

```{r screen_manhattan, fig.cap = "Manhattan-style plot of −log10(p) for Trait1. Known QTN positions are highlighted in red."}
# Known QTN positions for Trait1
qtn1 <- c("SNP003", "SNP015", "SNP025")

screen_plot <- screen_stats[!is.na(screen_stats$P.value), ]
screen_plot$neg_log_p <- -log10(screen_plot$P.value)
screen_plot$is_qtn    <- screen_plot$SNP %in% qtn1

plot(
  seq_len(nrow(screen_plot)),
  screen_plot$neg_log_p,
  pch  = ifelse(screen_plot$is_qtn, 18, 16),
  col  = ifelse(screen_plot$is_qtn, "firebrick", "steelblue"),
  cex  = ifelse(screen_plot$is_qtn, 1.8, 1.0),
  xlab = "SNP index",
  ylab = expression(-log[10](p)),
  main = "Marginal screening — Trait1",
  las  = 1
)
abline(h = -log10(0.05), col = "darkorange", lty = 2)
legend("topright",
       legend = c("Non-QTN", "QTN (known)", "p = 0.05"),
       pch    = c(16, 18, NA),
       lty    = c(NA, NA, 2),
       col    = c("steelblue", "firebrick", "darkorange"),
       bty    = "n", pt.cex = c(1, 1.5, 1))
```

## Multi-trait screening

When `y` is a numeric matrix (samples × traits), `compute_screening_stats()`
returns a **named list** of per-trait `data.table`s:

```{r screen_multi}
pheno_matrix <- pheno_multi$phenotype  # 50 × 2 matrix

screen_list <- compute_screening_stats(
  geno_mat = geno_mat_filt,
  y        = pheno_matrix,
  covar    = pheno_multi$covariates,
  verbose  = FALSE
)

# Named list, one data.table per trait
names(screen_list)
head(screen_list$Trait1[order(screen_list$Trait1$P.value), ])
head(screen_list$Trait2[order(screen_list$Trait2$P.value), ])
```

```{r screen_compare, fig.cap = "Comparison of −log10(p) between Trait1 and Trait2. Known QTN for each trait are highlighted."}
qtn2 <- c("SNP006", "SNP028", "SNP020")

p1 <- -log10(screen_list$Trait1$P.value)
p2 <- -log10(screen_list$Trait2$P.value)
xlim_max <- max(p1, na.rm = TRUE) * 1.05
ylim_max <- max(p2, na.rm = TRUE) * 1.05

plot(p1, p2,
     pch  = 16,
     col  = "grey60",
     xlab = expression(Trait1~~-log[10](p)),
     ylab = expression(Trait2~~-log[10](p)),
     main = "Marginal p-values: Trait1 vs Trait2",
     xlim = c(0, xlim_max), ylim = c(0, ylim_max))

# Highlight QTN
points(p1[screen_list$Trait1$SNP %in% qtn1],
       p2[screen_list$Trait1$SNP %in% qtn1],
       pch = 17, col = "firebrick", cex = 1.6)
points(p1[screen_list$Trait2$SNP %in% qtn2],
       p2[screen_list$Trait2$SNP %in% qtn2],
       pch = 15, col = "darkorange", cex = 1.6)

abline(h = -log10(0.05), v = -log10(0.05),
       col = "steelblue", lty = 2)
legend("topright",
       legend = c("Background", "Trait1 QTN", "Trait2 QTN", "p = 0.05"),
       pch    = c(16, 17, 15, NA), lty = c(NA, NA, NA, 2),
       col    = c("grey60", "firebrick", "darkorange", "steelblue"),
       bty    = "n")
```

---

# Candidate SNP selection

## Selection modes

Three modes map screening statistics to a candidate set $\mathcal{C}$:

**Mode A — p-value threshold**

$$\mathcal{C}_A = \{i : p_i \le \tau_p\}$$

```{r cands_A}
cands_A <- select_candidate_snps(
  screen_stats,
  mode           = "A",
  pval_threshold = 0.05
)
cat("Mode A candidates:", length(cands_A), "\n")
cands_A
```

**Mode B — effect-size criteria**

$$\mathcal{C}_B = \{i : |z_i| \ge \tau_z \;\text{[AND/OR]}\; R^2_i \ge \tau_{\text{pve}}\}$$

```{r cands_B}
# OR logic: keep SNPs with large z OR meaningful PVE
cands_B <- select_candidate_snps(
  screen_stats,
  mode          = "B",
  z_threshold   = 2.0,
  pve_threshold = 0.05,
  logic         = "OR"
)
cat("Mode B (OR) candidates:", length(cands_B), "\n")

# AND logic: require both criteria simultaneously
cands_B_and <- select_candidate_snps(
  screen_stats,
  mode          = "B",
  z_threshold   = 2.0,
  pve_threshold = 0.05,
  logic         = "AND"
)
cat("Mode B (AND) candidates:", length(cands_B_and), "\n")
```

**Mode C — hybrid**

$$\mathcal{C}_C = \{i : p_i \le \tau_p \;\text{[AND/OR]}\; |z_i| \ge \tau_z\}$$

```{r cands_C}
cands_C <- select_candidate_snps(
  screen_stats,
  mode           = "C",
  pval_threshold = 0.05,
  z_threshold    = 1.5,
  logic          = "OR"
)
cat("Mode C (OR) candidates:", length(cands_C), "\n")
```

## Threshold guidance

| Mode | Use case | Key parameters |
|------|----------|----------------|
| A | Clear significance cutoff available (e.g., Bonferroni, permutation threshold) | `pval_threshold` |
| B | No significance cutoff; select by effect magnitude | `z_threshold`, `pve_threshold`, `logic` |
| C | Capture SNPs significant OR large-effect (comprehensive) | all three thresholds + `logic` |

For Mode A, a Bonferroni threshold for the example data (40 tests):

```{r bonferroni}
alpha_bonf <- 0.05 / nrow(screen_stats)
cat("Bonferroni threshold:", formatC(alpha_bonf, format = "e", digits = 3), "\n")

cands_bonf <- select_candidate_snps(
  screen_stats,
  mode           = "A",
  pval_threshold = alpha_bonf
)
cat("Candidates at Bonferroni:", length(cands_bonf), "—", cands_bonf, "\n")
```

---

# Important SNP expansion

Candidate SNPs are expanded into a protected set $\mathcal{I}$ in two layers:

**Layer 1 — positional window:**

$$\mathcal{W}(c) = \{i : \text{chr}(i) = \text{chr}(c),\; |\text{pos}(i) - \text{pos}(c)| \le w\}$$

where $w = w_{\text{kb}} \times 1000$ bp (default $w = 50\,000$ bp each side).

**Layer 2 — LD neighbours (optional):**

$$\mathcal{L}(c) = \{i \in \mathcal{W}(c) : r^2(i, c) \ge \tau_{\text{flag}}\}$$

The protected set is the union over all candidates:

$$\mathcal{I} = \bigcup_{c \in \mathcal{C}} \left[\mathcal{W}(c) \cup \mathcal{L}(c)\right]$$

```{r expand}
important <- expand_important_snps(
  candidate_snps       = cands_A,
  snp_info             = snp_info_filt,
  geno_mat             = geno_mat_filt,
  window_kb            = 50,
  include_ld_neighbors = TRUE,
  r2_flag              = 0.90,
  verbose              = FALSE
)

cat("Candidate SNPs:  ", length(cands_A),   "\n")
cat("Important SNPs:  ", length(important), "\n")
cat("Added by expansion:", length(important) - length(cands_A), "\n")
cat("\nImportant SNP set:\n")
print(sort(important))
```

The $r^2$ statistic is the squared Pearson correlation computed from the dosage
matrix with pairwise-complete observations:

$$r^2(i, j) = \left[\frac{\text{Cov}(g_i, g_j)}{\sqrt{\text{Var}(g_i)\, \text{Var}(g_j)}}\right]^2$$

```{r r2_demo}
# Direct access to the pairwise r² matrix for a SNP subset
block_a_snps <- paste0("SNP", sprintf("%03d", 1:8))
r2_block_a   <- compute_r2_subset(geno_mat_filt, block_a_snps)
round(r2_block_a, 2)
```

---

# Background LD pruning

All SNPs outside $\mathcal{I}$ form the background set
$\mathcal{B} = \Omega \setminus \mathcal{I}$. A greedy forward-selection
algorithm is applied chromosome-by-chromosome in genomic position order:

$$\text{for each SNP } i \text{ in } \mathcal{B}_{\text{chr}}:
\quad \text{discard } j > i \text{ if } r^2(i, j) \ge \tau_{\text{genome}}$$

The default genome-wide pruning threshold is $\tau_{\text{genome}} = 0.80$.

```{r background_prune}
remaining <- setdiff(snp_info_filt$SNP, important)
cat("Background SNPs to prune:", length(remaining), "\n")

bg_retained <- prune_background_snps(
  remaining_snps = remaining,
  snp_info       = snp_info_filt,
  geno_mat       = geno_mat_filt,
  r2_genome      = 0.80,
  verbose        = FALSE
)

cat("Background retained after pruning:", length(bg_retained), "\n")
```

The final panel combines both sets:

```{r final_panel}
final_snps <- unique(c(important, bg_retained))
cat("Total final panel:", length(final_snps), "SNPs\n")

# Are known QTN in the final panel?
cat("\nTrait1 QTN retained?", all(qtn1 %in% final_snps), "\n")
cat("QTN1 IDs:", paste(qtn1[qtn1 %in% final_snps], collapse = ", "), "\n")
```

---

# Input cleaning

Some variant callers (e.g. NGSEP) produce genotype files with malformed lines
-- rows where embedded delimiter characters cause the column count to exceed
the header. `clean_genotype_file()` removes such lines by streaming through
the file in 50 000-line chunks:

- For VCF files, comment lines (`#`) are passed through and the column count
  is established from the `#CHROM` header line.
- For numeric CSV and HapMap files, the first line is the header.
- The separator is detected automatically from the file extension
  (`vcf` / `hmp` -> tab, everything else -> comma).

```{r clean_standalone, eval = FALSE}
# Standalone use: clean before analysis
clean_genotype_file(
  file_in  = "raw_ngsep.vcf.gz",
  file_out = "cleaned.vcf.gz",
  verbose  = TRUE
)
```

Alternatively, pass `clean_malformed = TRUE` to `run_sldp()` and the cleaning
step is integrated automatically as part of step 1:

```{r clean_integrated, eval = FALSE}
res <- run_sldp(
  genotype_file   = "raw_ngsep.vcf.gz",
  phenotype_file  = "phenotype.csv",
  output_file     = "panel_pruned.csv",
  format          = "vcf",
  clean_malformed = TRUE,   # cleans any format: vcf, hapmap, numeric
  trait_col       = c("BL", "NBL"),
  mode            = "A",
  pval_threshold  = 0.05,
  verbose         = TRUE
)
```

The cleaned temporary file is deleted automatically after reading completes.
No manual steps are required.


# The `run_sldp()` pipeline

`run_sldp()` executes all steps above in a single end-to-end call. The user
provides two files (genotype and phenotype) and an output path; everything
else is handled internally.

## Step-by-step overview

| Step | Action | Key parameters |
|------|--------|----------------|
| 1 | Read genotype file (optionally clean malformed lines) | `format`, `clean_malformed` |
| 2 | Read and align phenotype | `sample_col`, `trait_col`, `covar_cols` |
| 3 | Select scale strategy | `scale_strategy`, `gds_dir`, `n_cores` |
| 4 | MAF filter | `maf_threshold` |
| 5 (optional) | High-LD pre-pruning | `preprune_large`, `preprune_r2` |
| 6 | Extract post-filter SNPs (GDS only) | — |
| 7 | Marginal screening + candidate selection | `mode`, `pval_threshold`, `z_threshold`, `pve_threshold`, `threshold_logic` |
| 8 | Important SNP expansion | `window_kb`, `include_ld_neighbors`, `r2_flag` |
| 9 | Background LD pruning | `r2_genome`, `slide_max_bp` |
| 10 | Merge important + retained background | — |
| 11 | Write output file | `output_format` |
| 12 | Write pruning report (optional) | `stats_output_file`, `summary_output_file` |

## Single-trait run (Mode A)

```{r run_single, eval = TRUE}
out_file <- tempfile(fileext = ".csv")

res_single <- run_sldp(
  genotype_file  = geno_file,
  phenotype_file = pheno_file,
  output_file    = out_file,
  sample_col     = "Sample",
  trait_col      = "Trait1",
  covar_cols     = c("PC1", "PC2"),
  format         = "numeric",
  output_format  = "numeric",
  mode           = "A",
  maf_threshold  = 0.05,
  preprune_large = TRUE,
  preprune_r2    = 0.99,
  pval_threshold = 0.05,
  window_kb      = 50,
  r2_flag        = 0.90,
  r2_genome      = 0.80,
  verbose        = FALSE
)

# Summary of the result
cat("Scale strategy used:", res_single$scale_strategy, "\n")
cat("Candidate SNPs:     ", length(res_single$candidate_snps), "\n")
cat("Important SNPs:     ", length(res_single$important_snps), "\n")
cat("Background retained:", length(res_single$background_retained), "\n")
cat("Final panel size:   ", nrow(res_single$final_snp_info), "\n")

# Confirm known Trait1 QTN are in the final panel
cat("\nTrait1 QTN in final panel?\n")
cat(paste(qtn1, ":", qtn1 %in% res_single$final_snp_info$SNP), "\n")
```

## Inspecting the result object

```{r result_inspect}
# Final SNP metadata (ordered by CHR, POS)
head(res_single$final_snp_info)

# Screening statistics
head(res_single$screening_stats[order(res_single$screening_stats$P.value), ])

# Step-by-step pruning counts
print(res_single$pruning_stats)
```

## Visualising the pruning cascade

```{r pruning_waterfall, fig.cap = "Step-by-step SNP count through the OptSLDP pipeline."}
ps <- res_single$pruning_stats
ps <- ps[ps$step != "input", ]

barplot(
  ps$n_after,
  names.arg = ps$step,
  col       = "steelblue",
  border    = "white",
  las       = 2,
  cex.names = 0.75,
  ylab      = "SNPs retained",
  main      = "Pruning cascade (Trait1, Mode A)",
  ylim      = c(0, max(ps$n_before) * 1.1)
)
```

## Single-trait run (Mode B)

```{r run_mode_B}
res_B <- run_sldp(
  genotype_file   = geno_file,
  phenotype_file  = pheno_file,
  output_file     = tempfile(fileext = ".csv"),
  trait_col       = "Trait1",
  covar_cols      = c("PC1", "PC2"),
  mode            = "B",
  z_threshold     = 2.0,
  pve_threshold   = 0.05,
  threshold_logic = "OR",
  verbose         = FALSE
)

cat("Mode B (OR) — Final panel:", nrow(res_B$final_snp_info), "SNPs\n")
cat("Candidates:", length(res_B$candidate_snps), "\n")
```

## Single-trait run (Mode C)

```{r run_mode_C}
res_C <- run_sldp(
  genotype_file   = geno_file,
  phenotype_file  = pheno_file,
  output_file     = tempfile(fileext = ".csv"),
  trait_col       = "Trait1",
  covar_cols      = c("PC1", "PC2"),
  mode            = "C",
  pval_threshold  = 0.05,
  z_threshold     = 1.5,
  threshold_logic = "OR",
  verbose         = FALSE
)

cat("Mode C (OR) — Final panel:", nrow(res_C$final_snp_info), "SNPs\n")
```

---

# Multi-trait analysis

When `trait_col` is a character vector, `run_sldp()` runs screening and
candidate selection independently for each trait and then takes the **union**
of all per-trait candidate sets:

$$\mathcal{C}^* = \bigcup_{k=1}^{K} \mathcal{C}^{(k)}$$

The union $\mathcal{C}^*$ is then expanded and the background is pruned once.
This guarantees that every SNP important for any of the requested traits is
retained in the final panel.

## Multi-trait run

```{r run_multi}
res_multi <- run_sldp(
  genotype_file  = geno_file,
  phenotype_file = pheno_file,
  output_file    = tempfile(fileext = ".csv"),
  sample_col     = "Sample",
  trait_col      = c("Trait1", "Trait2"),
  covar_cols     = c("PC1", "PC2"),
  mode           = "A",
  pval_threshold = 0.05,
  window_kb      = 50,
  r2_flag        = 0.90,
  r2_genome      = 0.80,
  verbose        = FALSE
)

cat("=== Multi-trait results ===\n")
cat("Trait1 candidates:   ", length(res_multi$candidate_snps_per_trait$Trait1), "\n")
cat("Trait2 candidates:   ", length(res_multi$candidate_snps_per_trait$Trait2), "\n")
cat("Union candidates:    ", length(res_multi$candidate_snps), "\n")
cat("Important SNPs:      ", length(res_multi$important_snps), "\n")
cat("Final panel:         ", nrow(res_multi$final_snp_info), "\n")
```

## Union protection in action

The key test is whether SNP020 — a Trait2-only QTN outside all LD blocks — is
retained even though it is never a candidate under Trait1:

```{r union_test}
# SNP020 is a Trait2 QTN but not a Trait1 QTN
cat("SNP020 in Trait1 candidates?",
    "SNP020" %in% res_multi$candidate_snps_per_trait$Trait1, "\n")
cat("SNP020 in Trait2 candidates?",
    "SNP020" %in% res_multi$candidate_snps_per_trait$Trait2, "\n")
cat("SNP020 in union candidates? ",
    "SNP020" %in% res_multi$candidate_snps, "\n")
cat("SNP020 in final panel?      ",
    "SNP020" %in% res_multi$final_snp_info$SNP, "\n")
```

## Comparing single-trait and multi-trait panels

```{r compare_panels}
snps_only_t1  <- setdiff(res_single$final_snp_info$SNP,
                          res_multi$final_snp_info$SNP)
snps_added_mt <- setdiff(res_multi$final_snp_info$SNP,
                          res_single$final_snp_info$SNP)

cat("SNPs in single-trait panel:        ", nrow(res_single$final_snp_info), "\n")
cat("SNPs in multi-trait panel:         ", nrow(res_multi$final_snp_info),  "\n")
cat("SNPs added by multi-trait union:   ", length(snps_added_mt), "\n")
if (length(snps_added_mt)) cat("  Added:", snps_added_mt, "\n")

# Confirm all known QTN for both traits are in the multi-trait panel
all_qtn <- union(qtn1, c("SNP006", "SNP028", "SNP020"))
cat("\nAll QTN in multi-trait panel?",
    all(all_qtn %in% res_multi$final_snp_info$SNP), "\n")
```

---

# Scale strategies for large panels

`OptSLDP` selects a scale strategy automatically based on SNP count after
MAF filtering. The strategy governs all LD computations in the pipeline.

| Strategy | Default trigger | LD backend |
|----------|----------------|-----------|
| `in_memory` | $n \le 200\,000$ | `cor()` on full matrix in RAM |
| `chunked` | $200\,000 < n \le 2\,000\,000$ | `cor()` per chromosome |
| `gds` | $n > 2\,000\,000$ | `snpgdsLDMat()` from disk (requires SNPRelate) |

Adjust the thresholds for your hardware:

```{r scale_options, eval = FALSE}
options(
  optsldp.thresh_small  = 1e5,   # below this: in_memory
  optsldp.thresh_medium = 1e6    # below this: chunked; above: gds
)
```

Override the strategy explicitly:

```{r scale_override, eval = FALSE}
# Force GDS strategy for a large WGS panel
res_wgs <- run_sldp(
  genotype_file  = "wgs_panel.vcf.gz",
  phenotype_file = "phenotype.csv",
  output_file    = "wgs_pruned.csv",
  format         = "vcf",
  scale_strategy = "gds",
  gds_dir        = "/scratch/optsldp_gds",
  n_cores        = 16,
  trait_col      = "YLD",
  mode           = "A",
  pval_threshold = 1e-4
)

# For GDS runs, final_geno_mat is NULL — extract explicitly if needed
# (the output CSV is always written)
geno_final <- extract_final_geno(
  gds_path   = file.path("/scratch/optsldp_gds", "sldp_main.gds"),
  snp_ids    = res_wgs$final_snp_info$SNP
)
```

The `gds` strategy writes the genotype matrix to a binary SNPRelate file once,
then all LD queries stream from disk in blocks. All GDS file handles are
tracked in a package-level registry (`.sldp_gds_env`) and closed automatically
on pipeline exit or on error via `on.exit(.close_all_gds_handles())`.

---

# Writing output files

## Numeric dosage output

```{r write_numeric, eval = TRUE}
tmp_csv <- tempfile(fileext = ".csv")

write_numeric_genotype(
  snp_info = res_single$final_snp_info,
  geno_mat = res_single$final_geno_mat,
  file     = tmp_csv
)

# Verify
out_check <- read.csv(tmp_csv, check.names = FALSE)
cat("Output rows (SNPs):", nrow(out_check), "\n")
cat("Output cols (meta + samples):", ncol(out_check), "\n")
head(out_check[, 1:7])
```

## HapMap output

```{r write_hapmap, eval = TRUE}
tmp_hmp <- tempfile(fileext = ".hmp.txt")

write_hapmap_genotype(
  snp_info = res_single$final_snp_info,
  geno_mat = res_single$final_geno_mat,
  file     = tmp_hmp
)

# Verify
hmp_check <- read.table(tmp_hmp, header = TRUE, sep = "\t",
                         check.names = FALSE, comment.char = "")
cat("HapMap output rows:", nrow(hmp_check), "\n")
head(hmp_check[, 1:8])
```

## Pruning statistics report

```{r write_report, eval = TRUE}
stats_csv <- tempfile(fileext = ".csv")
summ_txt  <- tempfile(fileext = ".txt")

write_pruning_report(
  pruning_stats = res_single$pruning_stats,
  stats_file    = stats_csv,
  summary_file  = summ_txt
)

# Read the CSV report
report_df <- read.csv(stats_csv)
print(report_df)

# Read the plain-text summary
cat(readLines(summ_txt), sep = "\n")
```

---

# Memory and performance notes

## Chunked genotype reading

For large files (> `chunk_threshold` rows, default 200 000) the two-pass
strategy reduces peak RAM from $2 \times \text{file size}$ (fread +
as.matrix) to approximately one chunk worth of doubles:

| Panel size | Samples | Naive peak RAM | Chunked peak RAM |
|-----------|---------|----------------|-----------------|
| 500 K SNPs | 3 000 | ~12 GB | ~400 MB |
| 2 M SNPs | 3 000 | ~48 GB | ~400 MB |

The chunk size and threshold are tunable per call:

```{r chunk_params, eval = FALSE}
geno_obj <- read_numeric_genotype(
  file            = "large_panel.csv",
  chunk_rows      = 20000L,   # smaller chunks for very tight RAM budgets
  chunk_threshold = 100000L   # start chunking earlier
)
```

## Explicit garbage collection

`gc(FALSE)` is called in five locations to release allocator pressure
promptly:

1. After each genotype chunk is filled into the pre-allocated matrix (numeric
   and HapMap readers).
2. After the single-pass `fread()` call in small-file paths.
3. After each chromosome's $r^2$ matrix in `preprune_high_ld()`.
4. After each chromosome's $r^2$ matrix in `prune_background_snps()`.

The `FALSE` argument suppresses the memory-report printout; the call is
transparent to the user.

## Parallelism

The `n_cores` argument is forwarded to SNPRelate functions in the GDS strategy
(`snpgdsLDMat`, `snpgdsPruneLD`, `snpgdsSNPRateFreq`). The screening step
uses vectorised matrix algebra (BLAS `tcrossprod`) for all SNPs simultaneously
rather than per-SNP `lm()` calls. When multiple traits are requested, traits
are screened in parallel using a FORK cluster on Linux. The C++ LD kernel
(RcppArmadillo) calls BLAS DGEMM which is automatically multi-threaded by the
system BLAS (OpenBLAS or MKL).

---

# Session information

```{r sessioninfo}
sessionInfo()
```

---

# References

Zhu D, Zhao Y, Zhang R, Wu H, Cai G, Wu Z, Wang Y, Hu X (2023). Genomic
prediction based on selective linkage disequilibrium pruning of low-coverage
whole-genome sequence variants in a pure Duroc population. *Genetics Selection
Evolution*, **55**(1), 72. <https://doi.org/10.1186/s12711-023-00843-w>

Akohoue F, Herrera CC, Balanta SJC, et al. (2026). Enhancing genomic
prediction ability of blast resistance using genome-wide association
study-derived marker weights in two rice (*Oryza sativa* L.) populations.
*Theoretical and Applied Genetics*, **139**, 52.
<https://doi.org/10.1007/s00122-026-05159-z>
